---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
###             by Rashid Chohan

```{r setoptions, echo=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

```

```{r}
#Loading Libraries

library("ggplot2")      #For plotting
library(lattice)        #To plot panel graph

```

## PART I

### 1. Loading and preprocessing the data

    Loading Activity data from Activity-Data file
```{r}
  actdata <- read.csv("activity.csv")
```
  
### 2. Process/transform the data (if necessary) into a format suitable for your analysis

    No pre-processing of data is performed because we can make progress!

***
##Part II : 

### What is mean total number of steps taken per day?


```{r}
# Calculating total number of steps per day, using tapply
  dailySteps <- tapply(actdata$steps, actdata$date, sum, na.rm=TRUE)
```

###    1. Make a histogram of the total number of steps taken each day

```{r , fig.height=4,fig.cap="Histogram of Total Number of Steps per day"}
    # Display histogram with width=500 steps
    qplot(dailySteps,binwidth=500)
```

###    2. Calculate and report the **Mean** and **Median** total number of steps taken per day

***
```{r}
# Calculating Mean and Median of Daily Steps
    meanDailySteps = mean(dailySteps)
    medianDailySteps = median(dailySteps)
```    

### For Total Number of steps taken per day:
###      **Mean** = **`r meanDailySteps`**  and 
###      **Median** = **`r medianDailySteps`**


***
##Part III

### What is the average daily activity pattern?
```{r}
    meanStepsPerInterval <- aggregate(x=list(actdata$steps), by=list(actdata$interval), FUN=mean, na.rm=TRUE)
    colnames(meanStepsPerInterval) <- c("Interval", "Average_Steps")
```
#### 1. Make a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, fig.height=4}
###     Average Steps taken per Day (graph)

    #Graph showing Average Steps during 5-minute Time-Intervals, accross all days
    plot(meanStepsPerInterval$Interval,meanStepsPerInterval$Average_Steps,type="l",xlab="Interval", ylab="Average Steps", main="Average Steps taken per Day")
```

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
    peakInterval <- which.max(meanStepsPerInterval$Average_Steps)
    
    ## What's the time at this Interval of the day
    peakTime <- meanStepsPerInterval$Interval[peakInterval]
```
###    **Peak Time** is **`r peakTime`**, which corresponds to **Interval ** # **`r peakInterval`** for the day.
    
***
##Part IV:    Imputing missing values

### 1. Calculate and report the total number of missing values in the dataset
```{r}
    #Which cases i.e. rows are complete (without any NA values)
    cmpCases <- complete.cases(actdata)
    
    #Using the above computation, figure out total number of rows which are missing data (i.e. have NA values)
    missingCases <- sum(!cmpCases)
```

### **Total Number of Missing** values = **`r missingCases`**


#### 2. Devise a strategy for filling in all of the missing values in the dataset. 

#### IMPUTING STRATEGY :  
#### For each of the 5-minute Interval which has NA, we'll replace NA with **Mean Number of Steps (across all days)** .


#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
## Creating a copy Activity Data
        newActivityDataWithMeans <- actdata
        
        #Identify those intervals which has NA values
        IntervalsWithNAs <- which(is.na(actdata$steps))
```        

#### Assign **Average number of Steps** for each **interval with NA (value)**
```{r}
        newActivityDataWithMeans$steps[IntervalsWithNAs] <- meanStepsPerInterval$Average_Steps
```

###  4. Make a histogram of the total number of steps taken each day 
```{r}

        newdailySteps <- tapply(newActivityDataWithMeans$steps, newActivityDataWithMeans$date, sum, na.rm=TRUE)	

        ## Display histogram with width=500 steps
        qplot(newdailySteps,binwidth=500,main="Total Number of Steps (after replacing NAs)")
```

###     4.1) Calculate and report the **Mean** and **Median** total number of steps taken per day. 
```{r}
              ## Calculating Mean of Daily Steps
              meannewDailySteps = as.integer(mean(newdailySteps))
              
        
              ## Calculating Median of Daily Steps
              mediannewDailySteps = median(newdailySteps)
              
```
### After imputing:
###      Now the **Mean** of steps taken per day is  **`r meannewDailySteps`** 
###      and the **Median** of steps taken per day is **`r mediannewDailySteps`**


###     4.2) Do these values differ from the estimates from the first part of the assignment? 
###   Yes, before the Mean was **`r meanDailySteps`**  and now it is **`r meannewDailySteps`** 
###   While the Median has changed from **`r medianDailySteps`** to be **`r mediannewDailySteps`**

###     4.3) What is the impact of imputing missing data on the estimates of the total daily number of steps?
              
###     Yes, the after imputing the **Mean** has shifted to higher (i.e. more) steps per day. ####     This is as expected since we added steps, which were missing in the data!

***
##Part V

### Are there differences in activity patterns between weekdays and weekends?

### **Yes, there are differences between Weekdays and Weekend activity***

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
      #Define a vector for weekday or weekend distinction 
      dayofweekcolumn <- 
          sapply(X = newActivityDataWithMeans$date, 
              FUN = function(dayofweek) {
                      if (weekdays(as.Date(dayofweek)) %in% c("Saturday", "Sunday"))
                      { dayofweek <- "weekend" }
                      else { dayofweek <- "weekday" }
                    }
            )
      #Add this new column to the new dataset
      newActivityDataWithDayofWeek <- cbind(newActivityDataWithMeans, dayofweekcolumn)
      colnames(newActivityDataWithDayofWeek) <- c("steps","date","interval","dayofweek")

      #Creating a new vector for the Panel Plot graph
      meanStepsPerInterval_New <- aggregate(data=newActivityDataWithDayofWeek, steps ~ interval + dayofweek, FUN=mean)
      
      #Draw 
      xyplot(data=meanStepsPerInterval_New, steps ~ interval | dayofweek, type="l", layout=c(1,2), aspect="fill")
```
